\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-10-26 paper/paper.tex
More experiments; better default parameters.

% Names
\newcommand{\BWTmerge}{\textsf{BWT\nobreakdash-merge}}
\newcommand{\sdarray}{\textsf{sdarray}}
\newcommand{\CEU}{\textsf{CEU}}
\newcommand{\RS}{\textsf{RS}}

@@

\author{%
Jouni Sirén \\[0.5em]
{\small\begin{minipage}{\linewidth}\begin{center}
\begin{tabular}{c}
Wellcome Trust Sanger Institute \\
Wellcome Trust Genome Campus \\
Hinxton, Cambridge, CB10 1SA, UK \\
\url{jouni.siren@iki.fi}
\end{tabular}
\end{center}\end{minipage}}
}

@@

The dramatic decrease in the cost of DNA sequencing has flooded the world with \emph{sequence data}. The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are many other projects that are similar or greater in scale. For each sequenced genome, a sequencing machine produces a large number of \emph{reads} (short sequences) that cover the genome many times over. For a \~3~Gbp human genome, the total length of the reads is often \~100~Gbp or more.

@@

For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, so that no run continues from one block to the next one. For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$. Each of these seven increasing sequences is encoded using the \sdarray{} encoding \cite{Okanohara2007}. To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block. A \select{} query on the same sequence transforms position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block. Finally, we decompress the block to answer the query. \select{} queries and random accesses to the \BWT{} work in a similar way. There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a relatively short lexicographic range.

@@

We ran our experiments on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory. The system was running Ubuntu 12.04 on Linux kernel 3.2.0. All software was compiled with gcc/g++ version 4.9.2. We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB hard disk for temporary files.

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc}
\hline
\textbf{Dataset} & \textbf{Collection} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
\hline
\CEU             & NA12878             &       284 Gbp &          2.81G &               101 bp \\
\CEU             & NA12891             &       242 Gbp &          2.40G &               101 bp \\
\CEU             & NA12892             &       245 Gbp &          2.42G &               101 bp \\
\CEU             & Merged              &       771 Gbp &          7.63G &               101 bp \\
\hline
\RS              & AA                  &       433 Gbp &          4.69G &         73 or 100 bp \\
\RS              & TT                  &       432 Gbp &          4.68G &         73 or 100 bp \\
\RS              & AT                  &       275 Gbp &          2.98G &         73 or 100 bp \\
\RS              & TA                  &       355 Gbp &          3.84G &         73 or 100 bp \\
\RS              & Merged              &      1.49 Tbp &          16.2G &         73 or 100 bp \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

\item \RS{} is from the ReadServer project, which uses all low-coverage and exome data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases. We chose four of the files (AA, TT, AT, and TA).

@@

CEU: ropebwt + this vs ropebwt vs ropebwt2

RS: trade-offs with different parameters

Note: input/output formats, BWT-merge times include verification

@@

conclusions

sequence ordering: original order may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information

>>>>

% Names
\newcommand{\BWTmerge}{\textsf{BWT\nobreakdash-merge}}
\newcommand{\ropebwt}{\textsf{RopeBWT}}
\newcommand{\ropebwtii}{\textsf{RopeBWT2}}
\newcommand{\CEU}{\textsf{CEU}}
\newcommand{\RS}{\textsf{RS}}

@@

\author{%
Jouni Sirén\thanks{This work was supported by the Wellcome Trust grant [098051].} \\[0.5em]
{\small\begin{minipage}{\linewidth}\begin{center}
\begin{tabular}{c}
Wellcome Trust Sanger Institute \\
Wellcome Trust Genome Campus \\
Hinxton, Cambridge, CB10 1SA, UK \\
\url{jouni.siren@iki.fi}
\end{tabular}
\end{center}\end{minipage}}
}

@@

The dramatic decrease in the cost of DNA sequencing has flooded the world with \emph{sequence data}. The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are many other projects that are similar or greater in scale. For each sequenced genome, a sequencing machine produces a large number of \emph{reads} (short sequences) that cover the genome many times over. For a 3~Gbp human genome, the total length of the reads is often 100~Gbp or more.

@@

For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, so that no run continues from one block to the next one. For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$. Each of these seven increasing sequences is encoded using the \emph{sdarray} encoding \cite{Okanohara2007}. To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block. A \select{} query on the same sequence transforms position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block. Finally, we decompress the block to answer the query. \select{} queries and random accesses to the \BWT{} work in a similar way. There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a relatively short lexicographic range.

@@

We ran our experiments on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory. The system was running Ubuntu 12.04 on Linux kernel 3.2.0. All software was compiled with gcc/g++ version 4.9.2. We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB hard disk for temporary files. \BWTmerge{} was allowed to use 32 threads, while the other \BWT{} construction tools were limited to 4 or 5 parallel threads by design. All reported merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers extracted from the human reference genome.

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc}
\hline
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp \\
     & NA12891 &  242 Gbp & 2.40G &       101 bp \\
     & NA12892 &  245 Gbp & 2.42G &       101 bp \\
     & Merged  &  771 Gbp & 7.63G &       101 bp \\
\hline
\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp \\
     & TT      &  432 Gbp & 4.68G & 73 or 100 bp \\
     & AT      &  275 Gbp & 2.98G & 73 or 100 bp \\
     & TA      &  355 Gbp & 3.84G & 73 or 100 bp \\
     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp \\
     & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

\item \RS{} is from the \emph{ReadServer project}, which uses all low-coverage and exome data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases.

@@

\smallbreak\noindent\textbf{Benchmarking.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the native format used by \BWTmerge. Then we merged the \BWT{}s (in the given order). We used 128~MB or 256~MB run buffers and 256~MB or 512~MB thread buffers. The number of merge buffers was 4 or 5 with 512~MB thread buffers and 5 or 6 with 256~MB thread buffers, so that the files on disk were always merged from either 8~GB or 16~GB of thread buffers.

\begin{figure}[t!]
\begin{center}
\includegraphics{benchmark.pdf}%
\hspace{-0.3in}%
\includegraphics{comparison.pdf}
\end{center}
\caption{Left: Time/space trade-offs for merging four \BWT{} files containing a total of 1.49~Tbp. Label rXbYmZ denotes $X$~MB run buffers, $Y$~MB threads buffers, and $Z$ merge buffers. The dashed line represents the total size of the last pair of \BWT{}s to be merged. Right: Time/space trade-offs for building the \BWT{} of the 771~Gbp \CEU{} dataset.}\label{fig:benchmark}
\end{figure}

The results of the benchmarks can be seen in Figure~\ref{fig:benchmark} (left). Merging took 31.4 to 35.7~hours and required 145 to 166~gigabytes of memory, depending on the parameter values. This corresponds to an average speed of 8.27 to 9.40~Mbp/s for inserting 1.06~Tbp into the initial file AA. As the total size of the last pair of \BWT{}s was 124.2~gigabytes, the memory overhead was 21.1 to 41.5~gigabytes. The temporary files required 287 to 306~gigabytes of disk space. For the further experiments, we chose 128\nobreakdash-megabyte run buffers and 256\nobreakdash-megabyte thread buffers. We use 5 merge buffers in the \emph{small} variant (memory overhead 21.1~GB) and 6 merge buffers in the \emph{fast} variant (overhead 30.8~GB).

\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared the performance of \BWTmerge{} to the \ropebwt{} \cite{Li2011-2013} implementation of the BCR algorithm \cite{Bauer2013} and to \ropebwtii{} \cite{Li2014a}. These two tools are the fastest existing \BWT{} construction tools for read collections on general hardware \cite{Li2014a}. We ran \ropebwt{} with parameters \texttt{-btORf -abcr} and \ropebwtii{} with parameters \texttt{-bRm10g}. Due to memory constraints, we used the smaller \CEU{} dataset.

We built the \BWT{} of the dataset in four ways. We concatenated all input files, and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}. We built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{} with small settings. Finally, we built the individual \BWT{}s using parallel \ropebwt{} jobs, and then merged them using \BWTmerge{} with fast settings. In each case, the tools were allowed to output the \BWT{} files in their preferred formats.

The results can be seen in Figure~\ref{fig:benchmark} (right).
% FIXME discussion of the results; note that ropebwt2 did not finish

\smallbreak\noindent\textbf{ReadServer.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files using the fast settings.
% FIXME sizes in native/sga format, merging time/space

@@

conclusions: fast, low memory overhead. because BWTs are in-memory data structures, they can be built on the same system they are going to be used

sequence ordering: original order may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information; ReadServer currently uses RLO
